---
title: "Datasets"
---

Use this guide to choose a generated or externally supplied dataset, check what
is required to obtain it, and continue with the appropriate result shape. The
functions below are representative; use the complete
[function reference](../reference/index.html) for the full catalogue and
argument details.

## Choose a dataset

| Choose this when you need... | Dataset family | Representative functions | Acquisition and scale |
| --- | --- | --- | --- |
| A simple manifold or curve with a controllable sample size | Basic synthetic data | `sphere()`, `ball()`, `helix()`, `swiss_roll()`, `s_curve()` | Generated locally; choose the number of rows. |
| Cluster and geometry stress tests from *How to Use t-SNE Effectively* | Distill t-SNE simulations | `gaussian_data()`, `two_clusters_data()`, `random_walk()`, `trefoil_data()` | Generated locally; choose the number of rows. |
| Paper-derived topology or hierarchy examples | PaCMAP and topological examples | `s_curve_hole()`, `curve2d()`, `synthetic_hierarchical_data()`, `taspheres()` | Generated locally; size is controlled by function arguments. |
| A fixed 3D point cloud for embedding experiments | Mammoth point clouds | `download_mammoth10k()`, `download_mammoth50k()` | Downloaded JSON; 10,000 or 50,000 rows. |
| Grayscale image classification data with familiar train/test splits | MNIST and relatives | `download_mnist()`, `download_fashion_mnist()`, `download_kuzushiji_mnist()`, `download_qmnist()` | Downloaded; 70,000 rows for MNIST, Fashion-MNIST, and Kuzushiji-MNIST, or 120,000 for QMNIST. |
| Small color images in ten classes | CIFAR-10 | `download_cifar10()`, `show_cifar()` | Downloaded; 60,000 images. The list result's pixel matrix uses about 0.69 GiB. |
| Stereo views that vary object, pose, and lighting | Small NORB | `download_norb_small()`, `show_norb_object()` | Downloaded; 48,600 image pairs. The complete list result's pixel matrix uses about 3.34 GiB. |
| Multiple poses of individual objects | COIL object images | `download_coil20()`, `download_coil100()`, `show_coil_object()` | Downloaded; 1,440 or 7,200 images. Requires the suggested `png` package. |
| Text documents for an embedding pipeline | 20 Newsgroups | `download_twenty_newsgroups()` | Downloaded; 18,846 documents. Preprocess with tools such as `tm` or `tidytext`. |
| A small face-image collection | Frey and Olivetti faces | `frey_faces()`, `olivetti_faces()` | Supplied by the suggested `RnavGraphImageData` package; 1,965 or 400 images. |
| The original data used in historical Isomap examples | Isomap datasets | `download_isomap_swiss_roll()`, `download_isomap_faces()` | Downloaded archived Matlab data; 20,000 or 698 rows. Requires `R.matlab`; faces also require `gzip` or `uncompress`. |

MNIST contains handwritten digits. Fashion-MNIST substitutes clothing images
and is designed as a harder drop-in benchmark; Kuzushiji-MNIST substitutes
cursive Japanese characters. QMNIST extends the MNIST test set to 60,000
digits, giving 120,000 rows in total.

## Generate a local dataset

Synthetic datasets need no download or optional package. Here, the seed makes
the sampled points reproducible.

```{r local-sphere}
library(snedata)

set.seed(2026)
sphere_data <- sphere(n = 200)

dim(sphere_data)
round(range(sqrt(rowSums(sphere_data[c("x", "y", "z")]^2))), 6)
```

The result has 200 rows, three coordinate columns, and a color column. The
coordinate norm is 1 for every row, confirming that the points lie on the
sphere's surface.

## Work with downloaded images

The MNIST-family, CIFAR-10, Small NORB, and COIL downloaders default to the
legacy wide data frame. Ask for `as = "list"` when you want the shared
image-result structure: its `data` matrix contains one image per row, while
`meta` contains one matching metadata row with lower-case names. Where a
dataset has training and testing sets, `meta$split` records that identity
explicitly. The
[canonical image-result reference](../reference/image-results.html) documents
the complete schema.

The following workflow requires network access and is not run while this
article is built.

```{r mnist-workflow, eval=FALSE}
mnist <- download_mnist(as = "list")

training_rows <- mnist$meta$split == "training"
mnist_training <- mnist$data[training_rows, ]
dim(mnist_training)

show_mnist_digit(mnist, which(training_rows)[1])
```

`mnist_training` has 60,000 rows and 784 pixel columns, and the final call
displays the first training digit. Fashion-MNIST and Kuzushiji-MNIST use the
same visualization helper; CIFAR-10, Small NORB, and COIL have the matching
helpers shown in the selection table.

## Sources and terms for externally supplied data

The MIT license for `snedata` covers the package code, not data obtained from
external sources. The table below records the statements available from the
cited source pages when they were checked on 2026-07-16. “No separate terms
identified” means that the cited page provides access or citation information
but does not state a dataset license; it should not be read as either permission
or a restriction. Users remain responsible for checking current upstream terms
for their intended use.

| Dataset family | Source and citation | Terms status |
| --- | --- | --- |
| MNIST | The package uses the [`fgnt/mnist` mirror](https://github.com/fgnt/mnist) of [Yann LeCun's original files](https://yann.lecun.com/exdb/mnist/). Cite LeCun, Bottou, Bengio, and Haffner (1998), *Gradient-Based Learning Applied to Document Recognition*. | No separate dataset terms identified on the original page or mirror. |
| QMNIST | [Yadav and Bottou's QMNIST repository](https://github.com/facebookresearch/qmnist) and their 2019 paper, *Cold Case: The Lost MNIST Digits*. | The upstream README states that QMNIST uses the repository's [BSD-style license](https://github.com/facebookresearch/qmnist/blob/main/LICENSE). |
| Fashion-MNIST | The [Fashion-MNIST repository](https://github.com/zalandoresearch/fashion-mnist) and Xiao, Rasul, and Vollgraf (2017), *Fashion-MNIST: a Novel Image Dataset for Benchmarking Machine Learning Algorithms*. | The official repository publishes an MIT license. |
| Kuzushiji-MNIST | The [Kuzushiji-MNIST repository](https://github.com/rois-codh/kmnist) and Clanuwat et al. (2018), *Deep Learning for Classical Japanese Literature*. | The upstream README licenses the dataset and repository under [CC BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/) and supplies the requested attribution. |
| CIFAR-10 | The [CIFAR-10 source page](https://cave.cs.toronto.edu/kriz/cifar.html) and Krizhevsky (2009), *Learning Multiple Layers of Features from Tiny Images*. | The source page requests citation and supplies archive checksums; no separate dataset license is stated there. |
| Small NORB | The [NORB source and terms page](https://cs.nyu.edu/~yann/data/norb-v1.0/) and LeCun, Huang, and Bottou (2004), *Learning Methods for Generic Object Recognition with Invariance to Pose and Lighting*. | The source permits research use, prohibits sale, and requests citation. |
| COIL-20 and COIL-100 | Columbia's [COIL-20](https://cave.cs.columbia.edu/repository/COIL-20) and [COIL-100](https://cave.cs.columbia.edu/repository/COIL-100) pages and the corresponding Nene, Nayar, and Murase (1996) technical reports. | The source pages provide download and citation information; no separate dataset license was identified on those pages. |
| 20 Newsgroups | Jason Rennie's [20 Newsgroups page](http://qwone.com/~jason/20Newsgroups/) provides the by-date archive used here; the collection is associated with Lang (1995), *Newsweeder: Learning to Filter Netnews*. | No separate terms were identified on the by-date source page. The [UCI record for the related original collection](https://kdd.ics.uci.edu/databases/20newsgroups/20newsgroups.data.html) permits free educational use with attribution. |
| Mammoth point clouds | The JSON files come from [Understanding UMAP](https://github.com/PAIR-code/understanding-umap), which describes their downsampling from the [Smithsonian woolly mammoth model](https://3d.si.edu/object/3d/mammuthus-primigenius-blumbach:341c96cd-f967-4540-8ed1-d3fc56d31f12). | The Smithsonian model is CC0; the Understanding UMAP repository is Apache-2.0 licensed. |
| Isomap Swiss roll and faces | Archived copies of the original Isomap assets, cited to Tenenbaum, de Silva, and Langford (2000), *A Global Geometric Framework for Nonlinear Dimensionality Reduction*. | No separate dataset terms were identified on the archived source pages. |
| Frey and Olivetti faces | Supplied by the suggested [`RnavGraphImageData`](https://cran.r-project.org/package=RnavGraphImageData) package, whose documentation traces both datasets to Sam Roweis's data page and credits Brendan Frey and AT&T Laboratories Cambridge respectively. | These are not downloaded or redistributed by `snedata`; no separate dataset terms were identified in the supplying package documentation. |
